---
title: ''
output:
  pdf_document: default
  html_document: default
  word_document: default
editor_options:
  markdown:
    wrap: 100
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# MA256 Lesson 2 - Guided Data Exploration (AY24-1)

## Getting Started

There are two things we do every time we start working in R Studio:

1.  Turn on the Tidyverse Package. Think of it like opening your toolbox so you can use the tools
    inside of it

```{r libraries}
library(tidyverse)
```

2.  Tell R Studio where to find the files you want to use and reads in the house data so we can
    begin exploring it.

```{r read data}
# setwd("C:/Users/PUT YOUR PATH TO THE CSV HERE/")
# data <- read_csv("L02_Redfin-Nashville.csv")

## OR reach out to a website where the data is hosted:
# data <- read_csv("https://raw.githubusercontent.com/jkstarling/MA256/main/data/L02_Redfin-Nashville.csv")
```

## Real Estate investing as a LT!

So you have just graduated from USMA, completed your BOLC, and have arrived at Fort Campbell, KY.
You would like to buy a house that is within your means but plan on buying one that is big enough to
rent out to some of your army buddies in order to make some additional cash. We want to look at the
return on investment (ROI) percentage, which is the ratio of the (monthly) cash flow divided by the
purchase price.

## Calculate Monthly House Payments

Let's create a new variable that estimates our total monthly house payment including mortgage,
taxes, insurance, and HOA fees for every home so we can think about what we can afford. To do this,
we'll need to make some assumptions:

1.  We'll put 20% down so 0.8\*PRICE = Loan Amount for our mortgage PMT calculations
2.  \$5k in closing costs for a 6% fixed APR on a 30 year primary residence
3.  We'll use TN average annual effective property tax rate of 0.64% (.0064) of PRICE
4.  We'll use TN average annual homeowner's insurance cost of \$1,233
5.  We'll use the BAH rate for Fort Campbell for a married O-1
6.  You want to charge \$650/month per bedroom

```{r parameters}
# Closing_cost <- XXXXX
# TN_eff_tax <- XXXXX
# TN_homeowner_ins_mon <- XXXXX
# BAH_Campbell_O1M_mon <- 1623
# MPR <- XXXXX  #Monthly Percentage Rate is the assumed Annual 3% Rate divided by 12 months
# n <-XXXXX
# Room_rent <- XXXXX
```

### Simplify the data

Let's simplify our dataset to only include variables we care about i.e. `address`, `city`, `price`,
`home`, `lot size`, `#BD`, `#BA`, and `year built`.

You must save the new dataset somewhere to use it, just like the `read_csv` function.

```{r}
# house_data <- data %>% select(XXXXX, XXXXX, XXXXX, ....)
```

### Missing data

The HOA/MONTH column in `data` has missing (NA) values. Any time a formula includes a variable with
a NA value, it results in an NA value. In this case, NA means there are no HOA fees so to fix this,
we need to replace all NA values with \$0 so we can use HOA/MONTH in mathematical formulas for total
monthly cost. We do this by creating a dataframe so we can use the `is.na()` function to replace all
NA values with 0:

```{r}
# df <- data.frame(XXXXX) #Creates dataframe using HOA/MONTH variable in "data"
# 
# df[is.na(XXXXX)] <- 0 #The is.na() function replaces all "NA" with a numerical 0
```

### Updating column names

If you open the dataframe `df`, the column name is not very appealing and it will be difficult to
refer to it in formulas. Let's rename it using the `colnames()` function to make our life easier:

```{r}
# colnames(df)[1] <- "XXXXX"  #We renamed the 1st column in our df dataframe to "HOA.MONTH"
```

We can now use the `mutate()` command to create a new variable in house_data that is equal to the df
dataframe variable HOA.MONTH values. NOTE: If we don't save the dataset with the new variable
somewhere, we won't actually be able to use it later, so we overwrote the original dataset with the
new one that includes the new variable:

```{r}
# house_data <- XXXXX %>% mutate(XXXXX) 
```

Notice that the column name is copied over exactly as it originally appeared. If the dataframe had
multiple columns, they would have all been copied over.

### Creating new columns

Create new columns that calculate the cost to buy, monthly payments, and rent. To calculate the
mortgage payment per month, use the following equation:

M = $P [ r (1 + r)^n] / [(1 + r)^n - 1]$.

M = Total monthly payment.

P = The total amount of your loan.

r = Your interest rate, as a monthly percentage.

n = The total amount of months in your timeline for paying off your mortgage.

```{r}
# house_data = house_data %>%
#   mutate(CostToBuy = XXXXX * XXXXX + XXXXX,
#          Monthly_PMT = XXXXX + XXXXX + XXXXX * PRICE / 12 +
#           (0.8  * XXXXX * (XXXXX*(1+XXXXX)^XXXXX))/((1+XXXXX)^XXXXX - 1),
#          Rent = (XXXXX-1) * XXXXX)
```

Again, overwrite original data so it includes the new variables (cash flow and ROI):

```{r}
# house_data = house_data %>% 
#   mutate(Cash_Flow = XXXXX - XXXXX + XXXXX, 
#          Annual_ROI = 12 * Cash_Flow/CostToBuy) 
```

Since the annual ROI calculation doesn't account for utilities, annual maintenance, vacancy, &
turnover expenses, let's only consider homes with an annual ROI \> 20% (0.2)

```{r}
# ROI_house_data = house_data %>% filter(Annual_ROI > XXXXX) 
```

#### Changing notation

NOTE: If your Annual ROI is in scientific notation, you may convert it using the following code to
make the output easier to interpret:

```{r}
# options(scipen=0)  #scipen=0 turns on scientific notation. scipen=999 turns it off
```

### Plotting ROI

One would think that the more money I risk, the greater my compensation/ROI should be. Let's compare
CostToBuy with Annual ROI to see if that's true. Since both are quantitative, a scatter plot is
ideal:

```{r}
# ROI_house_data %>%
#   ggplot(aes(x = XXXXX,y = XXXXX)) + XXXXX() +
#   labs(x = "Total Acquisition Cost",y = "Annual Return on Investment (%)",
#        title = "Acquisition Cost vs. Annual ROI")
```

How do you interpret this plot? It seems that cheaper homes have a higher ROI, but why do we think
they are cheaper? In stats, our primary goal is to try to understand the variability in data. Let's
do that now.

### A better, more informative, plot.

Let's dive into this data more deeply by adding a 3rd dimension (color) which reflects the number of
bedrooms:

```{r}
# ROI_house_data %>%
#   ggplot(aes(x = XXXXX,y = XXXXX, color = as.factor(XXXXX))) + 
#   geom_point() +
#   labs(x = "Total Acquisition Cost",y = "Annual Return on Investment (%)",
#        title = "Acquisition Cost vs. Annual ROI")  # + geom_smooth(method = "lm") 
```

We see a trend within categories based on the number of bedrooms, perhaps because the maximum rental
income is capped based on the number of bedrooms so the more expensive a 3BR home is, the less ROI.

Note: to overlay the plot with the best fitting least squares regression lines we must remove the
#hashtag.

### An investment strategy (using cashflow)

My investment strategy is to buy the home with the largest \# of BRs and the least acquisition cost
without sacrificing quality. There are other concerns as well, like proximity to universities, size
of bedrooms, number of bathrooms, etc. but the main consideration is potential income vs expense.

This is great analysis, but ROI is not everything. Perhaps we care more about how much cash is going
into our pocket each month rather than ROI. After all, they say cash is king, and accruing cash
faster will allow me to buy more homes sooner. Let's compare the number of bedrooms to the amount of
Cashflow for all homes with a positive ROI.

```{r}
# CF_house_data = house_data %>%
#   filter(Annual_ROI > XXXXX) 

# CF_house_data %>% 
#   ggplot(aes(x = XXXXX, y = XXXXX)) + XXXXX() +
#   labs(title  = "Comparative Boxplot of Cash Flow versus Number of Bedrooms", 
#        x = "Number of Bedrooms", y="Cash Flow (USD)")
```

When comparing the 1 & 2 BR boxplots, the median cash flow almost triples when you spend a little
extra for a 2BR. When comparing the 2 & 3 BR boxplots, your median cash flow almost doubles when you
spend a little extra for a 3BR. However spending extra for a 4BR does not seem to do much in regard
to median cash flow (maybe \$200 increase). Spending extra for a 5BR home seems to reduce the median
cash flow, but it is not uncommon to cash flow upwards of \$2500/mo which is significantly better
than what 4BR homes are offering. 6BR homes don't seem worth it at all. Bigger homes should cost
more when considering materials cost to build them, but the 6BR home has cash flows comparable to
that of a 2BR home.

### A different investment strategy (using ROI)

The jump from 3 to 4 BR and potential from 4 to 5 BR homes is most interesting/significant so let's
now compare ROI vs #BR's using the cash flow data in boxplot form:

```{r}
# CF_house_data %>% 
#   ggplot(aes(x = XXXXX, y = XXXXX)) + XXXXX()+
#   labs(title  = "Comparative Boxplot of Annual ROI versus Number of Bedrooms", 
#        x = "Number of Bedrooms", y="Annual ROI")
```

Notice your median and 75% quartile annual ROI are worse for a 4BR compared to a 3BR which means the
additional cash flow and ROI from a 4BR home does not justify its cost. You should purchase a 3BR
home unless you find a phenomenal deal on a 4BR or 5BR.


### A personal story - Additional BR ("I" = a former mathlete instructor)

I once bought a 3BR/4BA home valued at \$244k and invested \$44k over the next month to update the
home and convert 2 large rooms into 2 more bedrooms by adding closets. 2 months after completing the
renovations, the home appraised for \$345k and rented for \$3k/mo, cash flowing \$1700/mo profit.

Could we apply this same strategy by buying a 2, 2.5, or 3 BA home, and paying for an additional BR
to be built?

What would be the approximate value gain assuming the bedroom is 10 ft x 15 ft? We can figure this
out by comparing the average price of 2BR 1350-1550 sqft homes, 3BR 1500-1700 sqft homes, and 4BR
1650-1850 sqft homes to estimate the value gain of a bedroom addition project.

```{r}
# Addition_data = house_data %>%
#   filter(XXXXX)
```

We use & (ampersand) to filter on multiple conditions simultaneously (e.g. "AND"). We use \|
(vertical line above the enter key) to filter "OR" conditions

Now let's compare the average price for each BR/BA combination within the SQFT ranges depending on
size to determine the value of a 1 BR addition: NOTE that the order of the group_by function will
simplify or complicate analysis so when you want to analyze the benefit of a 1x bedroom addition,
having BEDS 2nd in the group_by() function makes it easier to compare.

```{r}
# Addition_data %>%
#   group_by(XXXXX, XXXXX) %>% #calculates mean(PRICE) for every combination of BEDS & BATHS
#   summarise(Average_Price = XXXXX) 
```

Based on this information, there are only 3x scenarios that could generate profit: - Converting a
2BA home from 3BR to 4BR has a potential average gain of \$28,025. - Converting a 2.5BA home from
2BR to 3BR has a potential average gain of \$95,638. - Converting a 3BA home from 3BR to 4BR has a
potential average gain of \$71,250.

When I had a bedroom addition completed, it cost me \$15,000 so all 3 of these scenarios seem
profitable, but if I could only pick one, the \$95,638 seems best.


### Another personal story - Additional BA (where "I" is the same former mathlete instructor)

I once owned a 4BR/2.5BA home where the bathroom situation was garbage. There was an owner's suite
on the 3rd floor with it's own bathroom, but 2 other BR's on the 3rd floor and 1 more bedroom in the
basement all shared the other full BA that was in the hall on the 3rd floor. I spent \$15k to
install a 3rd full bathroom in the basement to make the basement suite like an additional owner's
suite. My property has gained \$80k in value almost immediately and allowed me to profit an extra
\$600/mo in rent.

Could we apply this same strategy by buying a 2, 3, or 4 BR home, and paying for an additional full
bathroom to be built? Using the same code as before, but having BATHS 2nd in the group_by command
makes it very easy to analyze:

```{r}
# Addition_data %>%
#   group_by(XXXXX, XXXXX) %>% #calculates mean(PRICE) for every combination of BEDS & BATHS
#   summarise(Average_Price = XXXXX) 
```

Based on this information, there are only 5x scenarios that could generate profit: 1. Converting a
2BR home from 1BA to 2BA has a potential average gain of \$143,000. 2. Converting a 3BR home from
1BA to 3BA has a potential average gain of \$109,038. 3. Converting a 3BR home from 2BA to 3BA has a
potential average gain of \$77,812. 4. Converting a 4BR home from 2BA to 3BA has a potential average
gain of \$121,037. 5. Converting a 4BR home from 3BA to 4BA has a potential average gain of
\$119,000.

You can conduct a similar analysis to consider the addition of 1BD AND 1BA: - Converting a 3BD/1BA
to a 4BD/2BA has a potential average gain of \$137,063 - Converting a 3BD/2BA to a 4BD/3BA has a
potential average gain of \$149,062 - Converting a 3BD/3BA to a 4BD/4BA has a potential average gain
of \$190,250

Considering everything at once, since 3BR homes had the best ROI and CF for the price, AND have the
most average profit potential given 1BD/1BA addition, buying a 3/3 and upgrading it to a 4/4 seems
to be the optimal investment strategy in the short term and the long term. If I'm willing to learn
and do some of the work myself I could make almost \$200k profit while cash flowing, \$1500/mo in
rent.

### Your Assignment

1)  Run the code above again, but vary a few of the numbers in the `parameters` code chunk above.
    Does it change any of the analysis? Which parameter will have the largest affect on ROI and
    cashflow?

2)  Find your own Redfin and attempt to draw out your own insights. Conduct the same analysis we did
    above, and translate your data into meaningful information and insights to inform your our house
    purchase!Often, good data scientists take the code of others (appropriately cited!) and then
    modify to fit their needs/ purposes! To complete this lesson, use this markdown file as a guide
    to produce similar plots and visualizations However, the true value will come from the insights
    you find from the visualizations. The markets have certainly changed since this data was pulled
    from Redfin, and the underlying assumptions we made for that place and that time might not be
    the same where you want to live!

In one paragraph, describe your new home buying plan and justify your approach based on what you
found through the process above. Good luck!

### Helpful websites to hone your strategy:

<https://www.travel.dod.mil/Allowances/Basic-Allowance-for-Housing/BAH-Rate-Lookup/> for BAH Rates

google: "average homeowners insurance cost in (insert state here)"

google: current home loan rates

<https://www.rocketmortgage.com/learn/closing-costs> for Closing Cost information

Feel free to adjust your rent with justification!
