---
title: ''
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 100
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# MA256 Lesson 2 - R Studio/Markdown Tips (AY24-1)

## R Markdown

-   **rmd** files operate much like `R` files except that they use **chunks** to execute lines of
    code.
-   outside of the **chunks**, we can use `LaTex` functionality to make our documents more visually
    appealing (see below for more \`mathy' stuff).
-   I **DO NOT** require that you use `RMarkdown` files over `R Scripts`, but the daily products I
    provide will be using `RMarkdown`.

## Starting out

-   You should always start from a **clean** RMarkdown file.
    -   Open up a new file (or delete everything in an already existing file)

    -   Save it as something different (in a folder that makes sense)

    -   Sweep your **Global Environment** with the broom icon in the upper right.

## Shortcuts

-   `alt|enter`: runs current line
-   `ctrl|enter`: runs current line **and** tabs down to next line
-   `ctrl|alt|i`: inserts new code chunk
-   `ctrl|shift|c`: comments line of code (or selection)
-   `ctrl|-`: zoom out
-   `ctrl|shift|+` : zoom in

## R Basics

```{r}
x = 1 # assigns value to variable
y <- 2 # another way to do leftward assignment
3 -> z # rightward assignment (NOT GOOD PRACTICE)
rm(x) # remove a variable from the global environment
rm(y,z) # remove several variables
```

## Vectors

```{r}
# a vector in R (an ordered list)
list1 = c(1, 2, 3, 2)
list1
list2 = c("a", "b", "c", "a")
list2
class(list1)
class(list2)
```

## Data frames

```{r}
# R likes to operate on vectors
df <- data.frame(list1, list2)
```

```{r}
# indexing in r
df$list1
df$list2
df[1, ] # returns the first row
df[ ,1] # returns the first column
```

```{r}
# functions in r
# format:    function(arg1, arg2, arg3)
mean(df$list1)
 # to get help on a function
```

## Tidyverse

-   `%>%` : pipe operator
-   used in `tidyverse` package to simplify nested commands
-   `ctrl|shift|m`: inserts pipe operator

## Packages

-   A package is a collection of functions.
-   Packages only need to be installed once, but we run the `library(insert_package_name_here)` at
    the start of each file.

```{r}
# Installing Packages
#install.packages("tidyverse")  
library(tidyverse) 
```

## Pipe Operator

```{r}
# Two ways of doing the same thing:
df %>% filter(list1 > 1)
filter(df, list1 > 1)

# Notice how this function has the same output
```

```{r}
# %>% becomes useful for multiple operations on an object:

df %>% 
  filter(list1 > 1) %>% 
  summarize(mean = mean(list1))

summarize(filter(df,list1 > 1), mean = mean(list1))

# Which method is more intuitive?
```

## Navigation

-   You can jump to chunks and titles by clicking at the bottom of this document.
-   Math operators in R: <https://www.datamentor.io/r-programming/operator/>
-   Clear variables with broom icon in Environment Tab (top right window)
-   The Plots Tab (bottom right window) and Console (bottom left window) also have the broom option.
-   You can run current chunk by clicking the green arrow at the top right of the chunk, by using
    `ctrl + alt + c`, or `ctrl|shift|enter`.
-   You can run the entire document up until your current chunk by clicking the down arrow at the
    top right of chunk.

## More Stuff

```{r}
x = 5 
class(x) # x is numeric type variable
y = c(5,6,7) # saves the list of numbers to the variable x 
class(y)
y = c(5,2,"cat") # all numbers are converted to strings so that all list elements are the same type
class(y) # now y is a character (aka "string") type variable
y[1] # access the first entry of y.  "5" is a string (aka "character" type)
class(y[1])
x[1] # access the first entry of z.   5 is a numbe r (aka "numeric" type)
class(x[1])

y = c(5,6,7)

x+y
z = c(1,2,3)
x + z
x*y
```

```{r}
# Remove variables with rm() function
rm(x) # remove one variable
rm(y,z) # remove multiple variables 

# Getting help on a function
# ?rm   # ?function_name to get help on a function
# Google is also a great reference for troubleshooting R
# You can copy error messages into Google and often find a solution
```

## Mathy stuff

We can also use LaTeX code to provide some mathy symbols:

-   We can make Greek letters, such as $\beta, \gamma, \sigma,$ and $\lambda$.

-   I can make a fraction, $\frac{a}{b}$, using in-line symbols.

-   I can also make the square root of a fraction of squared and cubed terms using `math mode`:
    $$\sqrt{\frac{a^2}{b^3}}$$

(optional) For more examples for LaTeX mathematics take a look at these sites:
<https://en.wikibooks.org/wiki/LaTeX/Mathematics>

<https://www.cmor-faculty.rice.edu/~heinken/latex/symbols.pdf>


