---
title: ''
geometry: "left=2cm,right=2cm,top=1cm,bottom=2cm"
output:
  pdf_document:
    extra_dependencies: ["xcolor", "amsmath","multicol"]
  html_document: default
editor_options:
  markdown:
    wrap: 100
---

```{r setup, message=FALSE, echo=TRUE, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newcommand{\sol}[1]{\color{violet} #1 \color{black}}

<!-- \renewcommand{\sol}[1]{} -->


<!-- xcolor standard colors: 
black, blue, brown, cyan, darkgray, green, lime, magenta, olive, orange, red, purple, teal, violet, yellow, etc.  -->

```{r include=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)

houses <- read_csv("https://raw.githubusercontent.com/jkstarling/MA256/main/data/Houses.csv")

```


# MA256 Lesson 18 \& 19 - Multiple Variables and Interactions (Causality and Multiple Regression Supplement Ch3 \& 4)

Many people say that you can predict the cost of a home (in thousands of USD) based on its size (square footage) because bigger homes cost more to build, so I’d like for you to verify this claim using a random sample of homes around Lake Michigan (Houses.csv). 

## A1.)	Identify and classify the variables of interest.

Explanatory: \sol{Home size (in square feet) - Quantitative}

Response:  \sol{Price (in thousands of USD) - Quantitative}

## A2.)	Define the coefficient symbols and interpret what they represent (in context) given the simple linear regression equation in the form of $y = \beta_0 + \beta_1*sqft$.

$\beta_0$ \sol{(Intercept Coefficient): the average price of a 0 sqft home (vacant lot). }

$\beta_1$ \sol{(Slope Coefficient): the population slope is the predicted increase in average price (in thousands of USD) for every one-square-foot increase in home size.}

3.	Express the null and alternate hypotheses in words & symbols.

$H_0:$ \sol{$\beta_1 = 0$. There is no association between the size of a home and its price.}

$H_a:$ \sol{$\beta_1 > 0$. There is a positive association between the size of a home and its price.} 



4.	Create a scatterplot of Home Size vs Price, and add a layer (`geom_smooth`) to include the best-fitting least squares regression line. Are the 4x validity conditions met for us to use the theory-based approach to assess strength of evidence? Use headers to explicitly state each of them, and then clarify if they are met.

```{r out.width="40%"}
houses %>% ggplot(aes(x = sqft, y = `price in thousands`)) + 
  geom_point() +
  geom_smooth(method = "lm") + 
  labs(x="Home Size in Square Feet", y = "Home Price in Thousands",
       title = "Home Price vs. Square Feet")
  
```

Linearity: \sol{The data seems to generally follow a linear trend.} 

Independence (New): \sol{It makes sense the home prices are independent from one another e.g. if my neighbor’s house has ugly paint inside that won’t influence my house price – or – if my neighbor’s house only has 2 bathrooms but mine has 4 that won’t affect my home price. }

Normality: \sol{Equal number of observations above (6) and below (7) the line. }

Equal Variance: \sol{There is an approximately equal variance throughout the entire domain, given flexibility due to small sample size. }




5.	Use the Course Guide example R Code for Lesson 18-19 as a template to create Coefficient Table and then write the complete equation for House_Model given the form of 
$y = \beta_0 + \beta_1 * sqft$.

```{r}
house.lm <- houses %>% lm(`price in thousands` ~ sqft, data = .)
summary(house.lm)
```

\sol{Price= -59.36870 + 0.21274*sqft}

6.	Interpret your slope and intercept coefficients:

Slope Coefficient: \sol{The increase in average price (in thousands of USD) for every 1 sqft increase in home size is 0.21274. }
Alternative Interpretation: \sol{The increase in average price for every 1 square foot increase in home size is $212.74 (0.21274 * 1,000 since price is in thousands).}


Intercept Coefficient: \sol{The average price (in thousands of USD) of a 0 square foot home is 
-59.36870. }
Alternative Interpretation: \sol{The average price of a 0 square foot home is -$59,368.70 
(-59.36870 * 1,000 since price is in thousands).}




7.	Would you be comfortable using your model to predict the price of a 0 sqft home (vacant lot)? Why or why not? 


\sol{Since positive numbers indicate money flowing from a buyer to a seller to purchase their property, and since the intercept coefficient is negative, it doesn’t make sense that someone would pay me $59,368.70 to “purchase” a vacant lot i.e. 0 square foot home. I am not comfortable using my model in this way. WE MUST ALWAYS USE CAUTION WHEN EXTRAPOLATING WITH OUR MODEL. }


Bonus: If we want to predict prices for vacant lots, what should we use to create a model?

\sol{If we wanted to predict the prices of vacant lots, perhaps we should only use vacant lots to create a new least squares regression model. }


8.	Calculate the 95% confidence interval for the slope statistic and provide an interpretation of the CI.

```{r}

```

\sol{The 2SD 95% CI = Statistic ± 2*SD. Since the slope coefficient (statistic) is 0.21274, and since its Standard Error (SE) is 0.06963, the 95% CI can be calculated as follows:\\
Lower Bound: 0.21274 – (2*0.06963) = 0.07348\\
Upper Bound: 0.21274 + (2*0.06963) = 0.352\\
95% CI: (0.07348, 0.352)\\
I am 95% confident that a 1 sqft increase in home size is associated with an average increase of 0.07348 thousands of USD to 0.352 thousands of USD in home price for homes around Lake Michigan.}

9.	Briefly explain why location i.e. lakefront (=1) or non-lakefront (=0), may be another explanatory variable. 

\sol{Lake houses are traditionally more expensive than non-lakefront houses. Lake impacts the response variables.}



10.	 Does lakefront affect price and sqft? Create a new scatterplot that is colored based on lakefront status. 
```{r}
houses %>% ggplot(aes(x = sqft, y = `price in thousands`, color = as.factor(lake))) +
  geom_point()
```


\sol{Based on the scatterplot results, it seems like even the smallest lakefront homes are more expensive than the largest non-lakefront homes that are twice the size (in sqft). }


11. We can adjust for the additional explanatory variable by including it in a multiple regression model. In this model, we estimate the price per square foot when we hold the location of the house constant.

$price_i = \beta_0 + \beta_1*sqft_i + \beta_2*lake_i$

where $price_i$ is the price of house $i$, $sqft_i$ is the size (sq ft) of house $i$, and $lake_i$ is 1 if house $i$ is lakefront and 0 if not lakefront. It may help to think of “$i$" as a counter from house #1 to #13

Create a new model with lake as an additional variable to create a scatter plot to view the new model. 

```{r}
house.lake.lm <- houses %>% lm(`price in thousands` ~ sqft + lake, data = .)
  
houses <- houses %>% mutate(price.predict = predict(house.lake.lm, newdata = .),
        	 residuals = `price in thousands` - price.predict)

houses %>% ggplot(aes(x = sqft, y = `price in thousands`, color = as.factor(lake))) +
  geom_line(aes(x=houses$sqft, y=house.lake.lm$fitted.values)) +
  geom_point(aes(y = `price in thousands`)) +
  labs(x="Home Size in Square Feet", y="Home Price in Thousands of USD", 
       		color="Lakefront Status", 
       		title="Home Price in Thousands of USD vs Size in Square Feet")

```



12.	Based on the slopes of the two lines, what assumption does this model make about the price per square foot? 

\sol{The slope for both the lakefront and non-lakefront homes is the same i.e. the price per square foot is the same regardless of whether a home is on the lake (no “interaction”). }




13.	How do we interpret $\beta_0,  \beta_1, \beta_2$ given the regression line equation in the form of $price_i = \beta_0 + \beta_1*sqft_i + \beta_2*lake_i$?

$\beta_0$: \sol{The price of a 0 square foot home (vacant lot) that is not on the lake.}

$\beta_1$: \sol{The change in average price of a home for every 1 square foot change in home size after adjusting for location (lakefront or not lakefront). }

$\beta_2$: \sol{The increase in average price because a home is lakefront after adjusting for size.}
	
	
14.	 What is our new regression model with lakefront status as an additional variable in the form of $price_i = \beta_0 + \beta_1*sqft_i + \beta_2*lake_i$? 

```{r}
summary(house.lake.lm)
```

\sol{$Price_i = -70.1821 + 0.1481*sqft_i + 331.2235*lake_i$}


15.	Calculate Prices of 2500 sqft houses on the lake and not on the lake. 
```{r}
predict(house.lake.lm, data.frame(sqft = 2500, lake = 1))
predict(house.lake.lm, data.frame(sqft = 2500, lake = 0))
```


\sol{2500 sqft Lakefront: -70.1821 + 0.1481*(2500) + 331.2235*(1) = \$631,291.40}

\sol{2500 sqft Non Lakefront: -70.1821 + 0.1481*(2500) + 331.2235*(0) = \$300,067.90}



16.	 How did the price per square foot (slope) change compared to when we held location constant?


\sol{The price per square foot decreased from \$212.74 (0.21274) to \$148.10 (0.1481) because the majority of the price difference was due to the lakefront premium. }


17.	 Which model predicts home price more accurately?

\sol{The first model had a coefficient of determination (R Squared) of 0.4591 so it accounted for 45.91\% of the variance in Price due to size. The second model has an R Squared of 0.9255 so it accounted for 92.55\% in home price due to size and location.}








