---
title: ''
geometry: "left=2cm,right=2cm,top=1cm,bottom=2cm"
output:
  pdf_document:
    extra_dependencies: ["xcolor", "amsmath","multicol"]
  html_document: default
editor_options:
  markdown:
    wrap: 100
---


## Lesson 18: ACFT Problem 
```{r setup, message=FALSE, echo=TRUE, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,   # avoid warnings and messages in the output
  message = FALSE,  # collapse all output into a single block
  tidy = FALSE,      # don't tidy our code-- assume we do it ourselves
  out.width = "50%"
)
```



```{r  include=FALSE}
# rm(list=ls())
# library(tidyverse)
# library(janitor)
```

```{r, include=FALSE}
# df <-  read_csv("https://raw.githubusercontent.com/acammack1234/ma206/main/acft_ioct.csv")
```

### Is there an association between the Standing Power Throw and Maximum Deadlift raw scores?

We are going to determine if an association exists between `spr_raw` and `mdl_raw` scores.

The relationship we are modeling looks like this:

$$\widehat{mdl} = \beta_0 + \beta_1 \cdot spr$$

1.  Provide the hypotheses in symbols and words:

\vfill

2.  Plot the association, including the regression line.

```{r}
# df %>%
#   ggplot(XXXXX
```

3.  Is there anything unusual about the data?

\vfill

4.  Remove the outlier using the `filter(spr_raw < 100)` command. How many observations were dropped?

```{r}
# df <- XXXXX 
```

5.  Create a new scatter plot with updated dataframe.

```{r}
# df %>%
#   ggplot(XXXXX
```

6.  Create a linear model to test your hypotheses.

```{r}
# lrmodel <- df %>%
#   XXXXX
# summary(lrmodel)
```

7.  Write out the linear model with the coefficients determined by the linear regression. 

\vfill


8.  Interpret the intercept term.

\vfill

\newpage

9.  Interpret the slope term.

\vfill

10. Conclusion of your hypothesis test at a 5% significance level.

\vfill

11. Interpret the R-squared value.

\vfill

12. Check the 4 Validity Conditions (L.I.N.E.). Are any not met?

```{r}
# lrmodel %>%
#   fortify(lrmodel$model) %>%
#   XXXXX
#   XXXXX
#   labs( x = "Predicted Values", 
#         y = "Residuals", 
#         title = "Residuals vs. Predicted Values")
# 
# lrmodel %>%
#   fortify(lrmodel$model) %>%
#   XXXXX
#   XXXXX
#   labs(x = "Order of Occurence", 
#        y = "Residuals", 
#        title = "Residuals in Order of Occurence")
# 
# lrmodel %>%
#   fortify(lrmodel$model) %>%
#   XXXXX
#   XXXXX
#   labs(x = "Residuals", 
#        title = "Histogram of Residuals")
```

\vfill

13. Can we determine causation? Why or why not.

\vfill


## Does adding the variable `sex` change the association between the Standing Power Throw (`spr_raw`) and Leg Tuck (`ltk_raw`) raw scores?

Now, we add `sex` as another variable to the model but also include it as an interaction with `spr_raw`.

    Our updated model looks like this:

$\widehat{mdl} = \beta_0 + \beta_1 \cdot spr + \beta_2 \cdot sexmale + \beta_3 (spr \cdot sexmale)$

14.  Plot the data **(including the interaction term)**, including the regression line.

```{r}
# df %>%
#   ggplot(XXXXX)
```

\newpage

15.  Does there appear to be a different relationship between `mdl_raw` and `spr_raw` based on `sex`?

\vfill

16.  Conduct a linear regression with the updated model.

```{r}
# lrmodel <- df %>%
#   lm(XXXXX)
# summary(lrmodel)
```

17.  Interpret the coefficient on the interaction term.

\vfill

18.  Is this significant at the 5% level?

\vfill

19.  Interpret the intercept term.

\vfill

20.  Interpret the base slope term for `spr_raw`.

\vfill

\newpage

21.  Using this model (even with no significance), calculate the following:

-   The `mdl_raw` for a male cadet with an `spr_raw` of 6m:

```{r}
#
```

-   The `mdl_raw` for a female cadet with an `spr_raw` of 6m:

```{r}
#
```



22.  Interpret the R-squared value.

\vfill

23. Check the 4 Validity Conditions (L.I.N.E.). Are any not met?

```{r}
# lrmodel %>%
#   fortify(lrmodel$model) %>%
#   XXXXXXXX 
#   XXXXXXXX
#   labs( x = "Predicted Values", 
#         y = "Residuals", 
#         title = "Residuals vs. Predicted Values")
# 
# lrmodel %>%
#   fortify(lrmodel$model) %>%
#   XXXXXXX
#   XXXXXXX
#   labs(x = "Order of Occurence", 
#        y = "Residuals", 
#        title = "Residuals in Order of Occurence")
# 
# lrmodel %>%
#   fortify(lrmodel$model) %>%
#   XXXXXXXXXX
#   XXXXXXXXXX
#   labs(x = "Residuals", 
#        title = "Histogram of Residuals")
```
